Parsing Postfix log files

### 2008-07-25 doc/logparser-implementation.tex
Lots of improvements to the efficiency section; lots more to come.

\section{Parser Implementation}

@@

        For efficiency the keywords are expanded and every rule's \regex{}
        is compiled before attempting to parse the log file --- otherwise
        each \regex{} would be recompiled each time it was used, resulting
        in a large, data dependent slowdown.  Efficiency concerns are
        discussed in \sectionref{parser efficiency}, with the impact of
        caching compiled \regexes{} covered in \sectionref{Caching each
        regex}.

>>>>

\section{Postfix Parser Implementation}

@@

        For efficiency the keywords are expanded and every rule's \regex{}
        is compiled before attempting to parse the log file --- otherwise
        each \regex{} would be recompiled each time it was used, resulting
        in a large, data dependent slowdown.  Efficiency concerns are
        discussed in \sectionref{parser efficiency}, with the impact of
        caching compiled \regexes{} covered in \sectionref{Caching compiled
        regexes}.

### 2008-07-25 doc/logparser-introduction.tex
Improved parts of the efficiency section and inlined graphs. Fixed up some references; added $Id$ to front page.
Lots of improvements to the efficiency section; lots more to come.

Section~\ref{rules} discusses the parsing rules in detail, explaining the
purpose and usage of each field in a rule, referring to an example rule and
sample data it matches successfully against.  The pros and cons of
overlapping rules are considered, including techniques for detecting
unintentional overlaps.  Rule efficiency concerns are discussed, in
particular the optimisations used by the algorithm, utilising the graphs in
\sectionref{graphs} for illustrative purposes.  The section concludes with
a description of using the tools provided with the parser to generate new
rules (specifically the \regex{} in each rule) from unparsed log lines.

@@

Section~\ref{graphs} contains graphs illustrating the topics previously
discussed in \sectionref{rule efficiency} (rule efficiency), and explains
the irregularities observed in these graphs.

>>>>

Section~\ref{rules} discusses the parsing rules in detail, explaining the
purpose and usage of each field in a rule, referring to an example rule and
sample data it matches successfully against.  The pros and cons of
overlapping rules are considered, including techniques for detecting
unintentional overlaps.  Rule efficiency concerns are discussed, in
particular the optimisations used by the algorithm.  The section concludes
with a description of using the tools provided with the parser to generate
new rules (specifically the \regex{} in each rule) from unparsed log lines.

@@

### 2008-07-25 doc/logparser-results.tex
Lots of improvements to the efficiency section; lots more to come.

@@

@@

Parsing efficiency is an obvious concern when the parser routinely needs to
deal with 75 MB log files containing 300,000 log lines (generated daily on
a mail server handling mail for approximately 700 users --- large scale
mail servers would have much larger log files on a daily basis).  When
generating the timing data used in this section, \numberOFlogFILES{} log
files (totaling 10.08 GB, \numberOFlogLINEShuman{} log lines) were each
parsed 10 times, the results of the first first run discarded, and the
execution time for the remaining 9 runs averaged.  The first run is
discarded for two reasons:

@@

@@

An important property of a parser is how execution time scales relative to
input size: does it scale linearly, polynomially, or exponentially?
\Graphref{execution time vs file size vs number of lines graph} shows the
execution time in seconds, file size in MB, and tens of thousands of log
lines per log file.  All three lines run roughly in parallel, giving the
impression that the algorithm scales linearly with input size.  This
impression is borne out by \graphref{execution time vs file size vs number
lines factor} which plots both the ratio of file size vs execution time,
and the ratio of number of log lines vs execution time (higher is better in
both cases); \tableref{execution time vs file size vs number lines factor
table} shows the ratios for different groups of log files.  As the
reader can see the ratios are quite tightly banded, showing that the
algorithm scales linearly: the much larger log files between points 60 and
70 on the X axis in \graphref{execution time vs file size vs number of
lines graph} actually cause the ratio to increase (i.e.\ improve), rather
than decrease.  The strange behaviour where larger log files are parsed
more efficiently is explained fully in \sectionref{Why are there dips in
the graphs?}.

@@

@@

\Graphref{percentage increase of shuffled over normal} shows the percentage
increase of shuffled over normal, with the percentage increase of reverse
over normal following in \graphref{percentage increase of reversed over
normal}; the mean increases for different groupings of log files are given
in \tableref{Execution time increase for different rule orderings}.
Overall this optimisation provides a modest but worthwhile performance
increase of approximately 10\%, for a small investment in time and
programming.

\label{rule ordering graphs}

XXX SHOW NORMAL AS A REDUCTION FROM SHUFFLE, AND REVERSE AS AN INCREASE
OVER SHUFFLE\@.

\showgraph{build/plot-normal-shuffle-factor}{Percentage increase of
shuffled over normal}{percentage increase of shuffled over normal}

\showgraph{build/plot-normal-reverse-factor}{Percentage increase of
reversed over normal}{percentage increase of reversed over normal}

\showtable{build/stats-normal-shuffle-reverse-include}{Execution time
increase for different rule orderings}{Execution time increase for
different rule orderings}

\subsubsection{Caching each regex}

\label{Caching each regex}

Perl compiles the original \regex{} into an internal representation,
optimising the \regex{} to improve the speed of matching, but this
compilation and optimisation takes CPU time; far more CPU time, in fact,
than the actual matching takes.  Perl automatically caches static
\regexes{}, but dynamic \regexes{} need to be explicitly compiled and
cached.  \Graphref{normal regex vs discard regex} shows execution
times with and without caching the \regex{}.  Caching the compiled
\regexes{} is obviously far more efficient; \graphref{normal regex
vs discarded regex factor} shows the percentage execution time increase
when not caching each \regex{}.

Caching the compiled \regexes{} is quite simple, and is the single most
effective optimisation implemented in the parser.

XXX ADD A TABLE\@.

@@

\label{graphs}

Graphs are an excellent means of displaying data, transforming a
meaningless stream of numbers into an easily comprehensible form, where
anomalies and patterns are immediately obvious.  The
graphs in the first section cover parser scalability, demonstrating that
performance scales linearly with input size.  The impact of rule ordering
is shown in \sectionref{rule ordering graphs}, and the anomalous dips and
peaks apparent in some graphs are explained.  The third group of graphs
vividly shows the huge impact that caching compiled \regexes{} has on
parser performance.  Miscellaneous graphs are presented in the final group
of graphs; these graphs are referenced from various places in the test.
This section concludes with a breakdown of the rule hits accumulated during
a single test run of the parser.

\subsubsection{Why are there dips in the graphs?}
\label{Why are there dips in the graphs?}

The dips at log files 22 and 62--68 correspond to peaks in log file size in
\graphref{execution time vs file size vs number of lines graph},
and peaks in \graphref{execution time vs file size vs number lines
factor} (where a peak means that more lines are processed per second, i.e.\
performance is better).  The explanation for this took some time to arrive
at, but it turns out to be reasonably simple.  The large log files in
question were caused by a mail forwarding loop, where the distribution of
log lines is quite different to normal, resulting in different performance
characteristics.

The mail loop was set up by a user modifying his mail forwarding to:
\newline \tab{}\texttt{$\backslash$username, username@domain} \newline This
instructs Postfix to deliver the mail to the local user, and also forward
it to the remote address; this is generally not a problem except that the
remote address in this case is the address the mail was originally sent to,
creating an infinite loop.  To prevent this happening Postfix examines the
Delivered-To header in the mail, and if the mail has already been delivered
to the current address it is bounced back to the sender with the error
message \texttt{mail forwarding loop for username@domain}.  Ordinarily this
works well, but unfortunately in this case the user noticed they had not
received any mail in a while and opted to send a test mail to themselves,
causing a loop not caught by Postfix:

    \item Postfix delivers the mail to the local mailbox and
        username@domain, as instructed by the user's forwarding
        instructions. The forwarded mail has a
        \texttt{Delivered-To:~username@domain} header added, and the
        envelope sender address is username@domain.  Log lines are added by
        \daemon{local}, \daemon{qmgr} (twice), \daemon{cleanup} and finally
        \daemon{pickup}.

    \item Postfix accepts the mail for username@domain, but while
        delivering it notices that the \texttt{Delivered-To} header already
        contains the address it is currently delivering to, and therefore
        sends a bounce notification with sender address \texttt{<>} to
        the original sender: username@domain.  Log lines are added by
        \daemon{local}, \daemon{qmgr} (twice) and \daemon{cleanup}.

    \item Postfix accepts the bounce notification and delivers it to both
        the local mailbox and to username@domain, as instructed by the
        user's forwarding instructions.  A
        \texttt{Delivered-To:~username@domain} header is added to the
        forwarded bounce notification, which now has an envelope sender
        address of username@domain.  Log lines are added by \daemon{local},
        \daemon{qmgr} (twice), \daemon{cleanup}, and finally
        \daemon{pickup}.

    \item Postfix accepts the forwarded bounce notification but while
        delivering the mail it notices that the \texttt{Delivered-To}
        header already contains the address currently being delivered to,
        and sends a bounce notification to the sender: username@domain.
        Log lines are added by \daemon{local}, \daemon{qmgr} (twice) and
        \daemon{cleanup}.

The vast majority of log lines when a mail loop occurs are from Postfix
components which have a small number of rules associated with them, whereas
in general \daemon{smtpd} adds the majority of log lines, and also has the
highest number of rules.  \daemon{smtpd} log lines are distributed across
rules much more evenly than the log lines of \daemon{qmgr}, \daemon{local},
\daemon{cleanup} or \daemon{pickup}, so the average number of rules
required to parse a \daemon{smtpd} log line is much higher that the average
number required to parse other log lines.

These two characteristics combine to reduce the average number or rules
required to parse a log line when there is a mail loop, as shown by the
peaks in \graphref{execution time vs file size vs number lines factor}.
When the rule ordering is reversed the majority of log lines generated by a
mail loop will be parsed with very few rules, whereas without a mail loop
the majority of log lines require a large number of rules; this leads to a
noticeable drop in the average time required to parse a log line, as shown
in \graphref{percentage increase of reversed over normal}.  The number of
rules which need to be consulted when the ordering is shuffled varies
between the optimal and the pessimal, and the performance varies
proportionally.

The difference between log files with a mail loop and log files without can
be seen in \tableref{Execution time increase for different rule
orderings} showing the increases for the different rule orderings and
combinations of log files:

\subsubsection{Caching regexes}

\label{Caching regexes}

Two large dips can be seen in \graphref{normal regex vs discarded
regex factor} at log files 22 and 62--68, corresponding to the spikes in
log file size in graphs~\refwithpage{execution time vs file size vs number
of lines graph} and~\refwithpage{normal regex vs discard regex}.  The
reason for the anomalous log files has already been explained in
\sectionref{Why are there dips in the graphs?}.

\subsubsection{Rule hits}
\label{rule hits}

@@

\clearpage

\subsubsection{Miscellaneous graphs}

\label{Miscellaneous graphs}

\clearpage

The first collection of graphs are about
parser scalability, showing the linear relationship between execution time
and input size.  \sectionref{rule ordering graphs} demonstrates the effect
of rule ordering on execution time, and the unexpected consequences of
specific inputs.  The necessity of caching compiled \regexes{} is attested
to by the third group of graphs, where the difference between caching and
not caching is staggering.  The penultimate section contains a breakdown of
the rule hits accumulated during a single test run of the parser.
Miscellaneous graphs expected to be useful are collected in the final
section.

>>>>

XXX NEED AN INTRODUCTION\@.

@@

XXX CHANGE normal TO optimal IN ORDERING STUFF, INCLUDING IN GRAPH CAPTIONS
AND AXES LABELS\@.

@@

XXX THIS SECTION NEEDS TO BE REWRITTEN\@.

Parsing efficiency is an obvious concern when the parser routinely needs to
parse very large log files.  The server which generated the log files used
in testing this parser accepts approximately 10,000 mails for 700 users a
day, and on average it produces 75MB log files containing 300,000 log lines
--- large scale mail servers would have much larger log files on a daily
basis.

When generating the timing data used in this section, \numberOFlogFILES{}
log files (totaling 10.08 GB, \numberOFlogLINEShuman{} log lines) were each
parsed 10 times, the results of the first first run discarded, and the
execution time for the remaining 9 runs averaged.  The results from the
first run are discarded for two reasons:

@@

\newpage

@@

An important property of a parser is how execution time scales relative to
input size: does it scale linearly, polynomially, or exponentially?
\Graphref{execution time vs file size vs number of lines graph} shows the
execution time in seconds, file size in MB, and tens of thousands of log
lines per log file.  All three lines run roughly in parallel, giving the
impression that the algorithm scales linearly with input size.  This
impression is borne out by \graphref{execution time vs file size vs number
lines factor} which plots both the ratio of file size vs execution time,
and the ratio of number of log lines vs execution time (higher is better in
both cases); \tableref{execution time vs file size vs number lines factor
table} shows the ratios for different groups of log files.  As the
reader can see the ratios are quite tightly banded, showing that the
algorithm scales linearly: the much larger log files between points 60 and
70 on the X axis in \graphref{execution time vs file size vs number of
lines graph} actually cause the ratio to increase (i.e.\ improve), rather
than decrease.  The strange behaviour where larger log files are parsed
more efficiently is explained fully in \sectionref{Explaining the peaks in
log file size}.

@@

\subsubsection{Explaining the peaks in log file size}

\label{Explaining the peaks in log file size}

\Graphref{execution time vs file size vs number of lines graph} shows log
file size varying between 30--100MB, except for log files 22 (200MB) and
62--68 (400--900MB).  These large log files were caused by a user
unintentionally creating a mail forwarding loop, resulting in a huge number
of mail deliveries and up to an order of magnitude more log lines than is
usual.  The user specified his mail forwarding as: \newline
\tab{}\texttt{$\backslash$username, username@domain} \newline This
instructs Postfix to deliver the mail to the local user, and also forward
it to the remote address; this is generally not a problem except that the
remote address in this case is the address the mail was originally sent to,
creating an infinite loop.  To prevent this happening Postfix examines the
Delivered-To header in the mail, and if the mail has already been delivered
to the current address it is bounced back to the sender with the error
message \texttt{mail forwarding loop for username@domain}.  Ordinarily this
works well, but unfortunately in this case the user noticed they had not
received any mail in a while and opted to send a test mail to themselves,
causing a loop not caught by Postfix:

    \item Postfix delivers the mail to the local mailbox and
        username@domain, as instructed by the user's forwarding
        instructions. The forwarded mail has a
        \texttt{Delivered-To:~username@domain} header added, and the
        envelope sender address is username@domain.

    \item Postfix accepts the mail for username@domain, but while
        delivering it notices that the \texttt{Delivered-To} header already
        contains the address it is currently delivering to, and therefore
        sends a bounce notification with sender address \texttt{<>} to
        the original sender: username@domain.

    \item Postfix accepts the bounce notification and delivers it to both
        the local mailbox and to username@domain, as instructed by the
        user's forwarding instructions.  A
        \texttt{Delivered-To:~username@domain} header is added to the
        forwarded bounce notification, which now has an envelope sender
        address of username@domain.

    \item Postfix accepts the forwarded bounce notification but while
        delivering the mail it notices that the \texttt{Delivered-To}
        header already contains the address currently being delivered to,
        and sends a bounce notification to the sender: username@domain.

The different distribution of log lines during a mail forwarding loop has
interesting effects on execution time, described in detail in
\sectionref{rule ordering for efficiency}, \sectionref{scalability as the
number of rules rises}, and \sectionref{Caching compiled regexes}.

@@

\Graphref{Execution time relative to
shuffled ordering graph} shows the execution time of normal and reverse orderings
relative to shuffled ordering; the mean increases for different groupings
of log files are given in \tableref{Execution time relative to shuffled
ordering table}.  Overall this optimisation provides a modest but worthwhile
performance increase of approximately 10\%, for a small investment in time
and programming.

\showgraph{build/plot-normal-and-reverse-relative-to-shuffle}{Execution
time of normal and reverse orderings relative to shuffled
ordering}{Execution time relative to shuffled ordering graph}

\showtable{build/stats-normal-shuffle-reverse-include}{Execution time
relative to shuffled ordering}{Execution time relative to shuffled ordering
table}

XXX IMPROVE THE REMAINDER OF THIS SECTION\@; CLEARLY SEPARATE THE
CHARACTERISTICS AND CLARIFY THEM\@.

Different rule orderings have less effect when parsing log files 22 and
62--68, due to the different distribution of log lines in those log files
(see \sectionref{Explaining the peaks in log file size}).  The vast
majority of log lines when a mail loop occurs are from Postfix components
which have a small number of rules associated with them, whereas in general
\daemon{smtpd} adds the majority of log lines, and also has the highest
number of rules.  \daemon{smtpd} log lines are distributed across rules
much more evenly than the log lines of \daemon{qmgr}, \daemon{local},
\daemon{cleanup} or \daemon{pickup}, so the average number of rules
required to parse a \daemon{smtpd} log line is much higher than the average
number required to parse other log lines.

These two characteristics combine to reduce the average number of rules
required to parse a log line when there is a mail loop.  The ratio of
number of lines versus execution time in \graphref{execution time vs file
size vs number lines factor} shows a noticeable improvement at log files 22
and 62--68, where a higher ratio means more log lines parsed per unit of
execution time.

When the rule ordering is reversed the majority of log lines generated by a
mail loop will be parsed with very few rules, whereas without a mail loop
the majority of log lines require a large number of rules; this leads to a
noticeable drop in the average time required to parse a log line, as shown
in \graphref{Execution time relative to shuffled ordering graph}.  The
number of rules which need to be consulted when the ordering is shuffled
varies between the optimal and the pessimal, and the performance varies
proportionally.

\subsubsection{Scalability as the number of rules rises}

\label{scalability as the number of rules rises}

XXX WRITE THIS SECTION\@.

\subsubsection{Caching compiled regexs}

\label{Caching compiled regexes}

Before Perl attempts to match a \regex{} against a piece of text, the
original \regex{} is compiled into an internal representation and optimised
to improve the speed of matching, but this compilation and optimisation
takes CPU time; far more CPU time, in fact, than the actual matching takes
in many cases.  Perl automatically caches static \regexes{} at compile
time, but \regexes{} loaded dynamically while the program need to be
explicitly compiled and cached.

Two large dips can be seen in \graphref{normal regex vs discarded
regex factor} at log files 22 and 62--68, corresponding to the spikes in
log file size in \graphref{execution time vs file size vs number
of lines graph} and \graphref{normal regex vs discard regex}.  The
reason for the anomalous log files has already been explained in
\sectionref{Explaining the peaks in log file size}.

XXX THIS IS AN AWFUL CAPTION\@.

This section began with parser scalability, showing the linear relationship
between execution time and input size.  Demonstrates the effect of rule
ordering on execution time, and the unexpected consequences of specific
inputs.  The necessity of caching compiled \regexes{} is attested to by the
third group of graphs, where the difference between caching and not caching
is staggering.  The penultimate section contains a breakdown of the rule
hits accumulated during a single test run of the parser.  Miscellaneous
graphs expected to be useful are collected in the final section.

@@

@@

\clearpage{}

### 2008-07-25 doc/logparser.tex
Lots of improvements to the efficiency section; lots more to come.

% Sort numbers where there are multiple citations.  does not appear to have
% any effect (probably clashes with varioref or hyperref), though it does
% reduce the space between numbers.
\usepackage{cite}

@@

\title{Parsing Postfix log files}
\author{John Tobin \\ School of Computer Science and Statistics \\
Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
\maketitle
\SVN$Id$
\SVNId{}

@@

\newpage

XXX NEW LAYOUT\@:

\begin{verbatim}
Introduction
Background
State of the Art Review
Parser Architecture
Postfix Parser Implementation
    Database Schema
    Rules
    Actions
    Complications
    Limitations and Improvements
Results - BETTER TITLE?
    Coverage
    Efficiency
Conclusion
References
Glossary
Acronyms
\end{verbatim}

>>>>

% Sort numbers where there are multiple citations.  Does not appear to have
% the expected effect (probably clashes with varioref or hyperref), though
% it does reduce the space between numbers.
\usepackage{cite}

@@

\title{Parsing Postfix log files}
\author{John Tobin \\ School of Computer Science and Statistics \\
Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
\maketitle
\SVN$Id$
\begin{center}\SVNId{}\end{center}

@@

XXX CHECK THE CAPTIONS ON ALL GRAPHS AND TABLES --- SOME ARE AWFUL\@.
